Persistent somatic symptoms are key to individual illness perception at one year after COVID-19 in a cross-sectional prospective cohort study

### 2023-02-09 paper/markdown/supplementary_material.Rmd
Second revision
Minor Reviewer Issues, duplicated references removed

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

@@

The CovILD longitudinal observation cohort (ClinicalTrials.gov: NCT04416100) includes adult ($\geq$ 18 year) patients with a symptomatic, PCR-confirmed SARS-CoV-2 infection. 
The follow-up visits were scheduled at two, three, six months and one year after COVID-19 diagnosis. 
In the current report, a subset of the original CovILD cohort was investigated, which displayed (1) persistent somatic symptoms (PSS) or (2) any abnormality in chest computed tomography (CT) or (4) any deficits in lung function testing (LFT) or (5) any cardiological abnormality in trans-thoracic echocardiography (TTE) at the one-year follow-up. 
The final analysis inclusion criterion was availability of complete Brief Illness Perception Questionnaire (IPQ) [@Broadbent2006] and the complete set of variables used in explorative data analysis, modeling and clustering (__Supplementary Table S\@ref(tab:tab-var-dict)__ and __S\@ref(tab:tab-modeling-vars)__).

The decision to analyze the study participant subset with COVID-19 sequlae was motivated by the design of the illness perception-measuring tool used in our report, the Brief Illness Perception Questionnaire (BIPQ), which was initially developed for and validated in somatic conditions [@Broadbent2006].

@@

The protocol of each follow-up visit included a survey of PSS, blood laboratory parameter assessment, LFT, TTE and chest CT as described in more detail before [@Sonnweber2020; @Sonnweber2022; @Luger2022]. 
Comorbidities and data on acute COVID-19 course (symptoms, laboratory parameters, functional tests and imaging) were recorded retrospectively at the two-month follow-up visit based on information provided by the participant and electronic patient records.

@@

The baseline demographic variables included sex (male/female), age at COVID-19 diagnosis, smoking history, body weight and height. 
The following pre-existing comorbidities were surveyed: obesity, metabolic disease, cardiovascular disease, hypertension, type II diabetes, hypercholesterolemia, gastrointestinal disease, malignancy, chronic kidney disease, immune deficiency, pulmonary and respiratory conditions. 
Among the surveyed comorbidities, asthma, COPD (chronic obstructive pulmonary disease), interstitial lung disease and other chronic pulmonary disorders were subsumed under the 'respiratory disease' variable included in the independent modeling variable set.
Study participants were stratified according to the acute COVID-19 severity as ambulatory (outpatient, WHO ordinal scale for clinical improvement 1 - 2), moderate (hospitalized at normal infection ward, no oxygen therapy, WHO 3 - 4) and severe COVID-19 survivors (hospitalized with oxygen therapy or mechanical ventilation or intensive care, WHO 5 - 7).

@@

LFT abnormality was defined as FVC (forced vital capacity) < 80% or FEV1 (forced expiratory volume in 1 second) < 80% or TLC (total lung capacity) < 80% or DLCO (diffusion lung capacity for carbon monoxide) < 80% of the predicted reference value or FEV1:FVC ratio < 70% of the predicted reference value [@Sonnweber2022]. 
CT images were screened for ground glass opacities (GGO), consolidations, bronchiectasis and reticulations according to the Fleischner Society glossary [@Hansell2008]. 
Chest CT abnormalities were scored separately for each lobe with the CT severity score (CTSS) [@Sonnweber2020; @Sonnweber2022; @Luger2022]: 0 - no abnormality, 1 - minimal (subtle GGO), 2 - mild (several GGO, subtle reticulation), 3 - moderate (multiple GGO, reticulation, small consolidation), 4 - severe (extensive GGO, consolidation, reticulation with distortion), 5 - massive (multiple findings, parenchymal destruction). 
The sum CTSS for all five lobes was used in the analysis. 
Any chest CT abnormality is defined as CTSS $\geq$ 1. 
The most frequent cardiological abnormality in TTE was low grade diastolic dysfunction. 
No reduced left ventricular ejection fraction was observed in the participants included in the current analysis. 
Frequencies of cardiopulmonary findings at the one-year follow-up are presented in __Supplementary Table S\@ref(tab:tab-cohort-fup)__ and __Supplementary Figure S\@ref(fig:fig-overlap)__.

@@

- __reduced physical performance__ assessed with the ECOG scale (Eastern Cooperative Oncology Group). Reduced physical performance was defined as ECOG $\geq$ 1
- __dyspnea__ rated with the mMRC scale (Modified Medical British Research Council). Dyspnea was defined as mMRC $\geq$ 1
- __cough__, self-reported, surveyed as a yes/no item
- __sleep problems__, self-reported, surveyed as a yes/no item
- __night sweating__, self-reported, surveyed as a yes/no item
- __hyposmia or anosmia__, self-reported, surveyed as a yes/no item
- __dermatological symptoms__, self-reported, surveyed as a yes/no item
- __gastrointestinal symptoms__, self-reported, surveyed as a yes/no item
- __hair loss__, self-reported, surveyed as a yes/no item
- __significant fatigue__, measured with the bimodal Chalder's Fatigue Scale (CFS). Significant fatigue was defined as bimodal CFS $\geq$ 4 [@Chalder1993; @Morriss1998]

@@

Since only few participants displayed values of ECOG $\geq$ 2 or mMRC $\geq$ 2, the dichotomous 'reduced performance (ECOG $\geq$ 1)' and 'dyspnea (mMRC $\geq$ 1)' independent variables were used in modeling instead of the full numeric scales.
Fatigue intensity was rated with the likert Chalder's Fatigue Scale (CFS) [@Chalder1993; @Morriss1998]. 
Exertional capacity was assessed by six-minute walking distance (SMWD) test performed according to the American Thoracic Society guidelines [@Crapo2012]. 
The reference SMWD values were obtained as described by Crapo and colleagues [@Crapo2012] and differences between the actual and reference values were calculated.
Frequencies of symptoms, ratings of fatigue and exertional capacity at the one-year follow-up are shown in __Supplementary Table S\@ref(tab:tab-cohort-fup)__, __Supplementary Figures S\@ref(fig:fig-sev-symptoms)__ and  __S\@ref(fig:fig-overlap)__.

@@

Laboratory parameters encompassing hemoglobin (total and glycated hemoglobin [HbA1c]), biomarkers of iron turnover (ferritin [FT], transferrin saturation [TF-Sat], hepcidin and soluble transferrin receptor [sTFR]), inflammation and coagulation (C-reactive protein [CRP] and D-dimer) and cardiovascular pathology (N-terminal pro-brain natriuretic peptide [NT-proBNP]) were determined routinely at the certified laboratory of the Central Institute for Medical and Chemical Diagnostic at the University Hospital in Innsbruck. 
Other markers of systemic inflammation, interleukin 6 (IL6) and procalcitonin (PCT), were within the normal ranges in the analysis collective. 
Values of laboratory parameters at the one-year follow-up are listed in __Supplementary Table S\@ref(tab:tab-cohort-biochemistry)__.

@@

Illness perception (IP) was scored with a German translation of the Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006]. 
Each of the 8 BIPQ items (questions Q1 - Q8) was scored in with an 11-point likert scale, the items Q3, Q4 and Q7 were conceptualized as negative items are were inverted prior to the analysis:

@@

The total IP score was calculated as an arithmetic sum of items Q1 - Q8. 
As shown by the results of factor analysis [@Bartlett1937] of the BIPQ tool items (__Supplementary Figure S\@ref(fig:fig-factor-analysis)__), two separate factors could be identified: the emotion/concern/consequences (items Q1, Q2, Q5, Q6, Q8) and the lacking control/coherence component (items Q3, Q4, Q7). 
For this reason, two IP component scores were calculated and analysed as well: the emotion/concern/consequences score defined as the Q1, Q2, Q5, Q6, Q8 item sum and the lacking control/coherence score defined as the Q3, Q4, Q7 item sum. 
Values of the total IP score, the emotion/concern/consequences and lacking control/coherence IP component scores as well as values of the BIPQ items recorded at the one-year follow-up are presented in __Table 2__.

@@

For descriptive statistic, correlation and statistical testing for differences between analysis groups the package _rstatix_ [@Kassambara2021] and the development package _ExDA_ were used (https://github.com/PiotrTymoszuk/ExDA) were employed.
Factor analysis was accomplished with the _stats_ package of base R and the development package _clustTools_ (https://github.com/PiotrTymoszuk/clustTools). 
The BIPQ score consistency was investigated with the _psych_ package [@Revelle2022].

For modeling and ANOVA, the development package _lmqc_ (https://github.com/PiotrTymoszuk/lmqc, backward elimination and two-way ANOVA) and the packages _glmnet_ (Elastic Net and LASSO) [@Friedman2010], _monomvn_ [@Gramacy2022] (Bayesian LASSO) and _caret_ [@Kuhn2008] were employed. 
Model quality control, cross-validation and fit statistics were computed with _caret_ and the development package _caretExtra_ (https://github.com/PiotrTymoszuk/caretExtra).

Clustering was accomplished with tools provided by the packages _cluster_ [@Schubert2019], _philentropy_ [@Drost2018], _factoextra_ [@Kassambara2020] and the development package _clustTools_ (https://github.com/PiotrTymoszuk/clustTools).

The _ggplot2_ [@Wickham2016] package were used for data visualization. 
Bar and violin plots were generated with _ExDA_. 
Quasi-proportional Venn diagrams were generated with the _nVennR_ package [@Perez-Silva2018]. 
Figures were constructed with _cowplot_ [@Wilke2019]. 
Tables were generated with _flextable_ [@Gohel2022]. 
The manuscript and supplementary material files were created with the _rmarkdown_ [@Allaire2022], _knitr_ [@Xie2022], _bookdown_ [@Xie2016] packages and _Affiliations_ Pandoc filter set (https://github.com/drdanholmes/Affiliations). 
For management of figures, tables and R code in the RMarkdown documents, the development package _figur_ (https://github.com/PiotrTymoszuk/figur) was used.

@@

Descriptive statistic: median with interquartile ranges for numeric variables, and percentages and counts for categorical variables were calculated with the `explore()` function (_ExDA_ package). 
Statistical significance of differences in frequency of categorical variable levels was assessed by $\chi^2$ test with Cramer V effect size statistic (`compare_variables()`, the _ExDA_ package). 
Normality was tested with Shapiro-Wilk test and quantile-quantile plots (`explore()`, _ExDA_ package).

Since multiple study variables were non-normally distributed, differences in numeric variables were assessed by non-parametric Mann-Whitney (effect size: r) or Kruskal-Wallis test (effect size: $\eta^2$), as appropriate. 
Group comparisons were done with the `compare_variables()` function (package _ExDA_).
Correlations were analyzed with Spearman test (`correlate_variables()` package _ExDA_) and visualized by fitting a second order trend (`geom_smooth()`, package _ggplot_, formula: $y \sim x + I(x^2)$). 
Statistical significance for changes in frequency of PSS in time was investigated with Cochran Q test (function `cochran_qtest()`, package _rstatix_) [@Kassambara2021].

@@

BIPQ dimensionality was assessed by maximum  likelihood factor analysis [@Bartlett1937] (`reduce_data(red_fun = 'fa')`, package _clustTools_, a wrapper around `factanal()` from _stats_) and visual investigation of the BIPQ item loadings (method `plot(type = 'loadings')` called for the factor analysis object). 
The factor analysis revealed presence of two components of the BIPQ tool: the emotion/concern/consequences and the lacking control/coherence factor (__Figure \@ref(fig:fig-factor-analysis)__).
Spearman's pair-wise correlations between the BIPQ items were calculated (function 'correlate_variables()', package _ExDA_) and visualized in a bubble plot. 
Due to multi-dimensionality of the BIPQ tool McDonald's $\omega$ was calculated with the function `omega()` from the package _psych_ [@Revelle2022; @McDonald1999] was used as a reliability measure for the global BIPQ and the emotion/concern/consequences and lacking control/coherence factors (__Supplementary Figure S\@ref(fig:fig-coherence)__).

@@

In multi-parameter modeling, a set of `r length(mod$variables)` independent variables was used (__Supplementary Table S\@ref(tab:tab-modeling-vars)__). 
To improve normality, FT and sTFR were transformed with the logarithm function, and TF-Sat and hepcidin were transformed with square root. 
Strongly non-normally or bimodally distributed numeric explanatory variables: 
number of comorbidities (`r levels(cov_data$clear_data$no_comorb_class) %>% paste(collapse = ', ')`), 
number of symptoms (`r levels(cov_data$clear_data$sympt_number_class) %>% paste(collapse = ', ')`), 
CTSS ('none': CTSS 0, 'mild': 1 - 5, 'moderate': 5 - 10, 'severe': 11+) [@Luger2022], 
NT-proBNP (`r levels(cov_data$clear_data$NTproBNP_class) %>% paste(collapse = ', ')`), 
D-Dimer (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`), 
CRP (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`) and 
HbA1c (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`) were stratified with clinically applicable cutoffs or cutoffs determined by visual analysis of the plots of modeling variable and the explanatory factor with fitted LOESS trends. 
For numeric explanatory variables, both first- and second-order terms were included in the model to account for possible non-linear relationship between the independent variable and response. 
During pre-processing, numeric variables (independent and responses) were normalized with mean centering (Z-scores).

@@

Three L1 regularized regression algorithms were utilized to model the total IP score and the IP component scores at one year after COVID-19: Elastic Net [@Zou2005], LASSO (least absolute shrinkage and selection operator) [@Tibshirani1996] and Bayesian LASSO [@Park2012]. 
The choice of penalized regression over canonical backward elimination or forward selection was motivated by highly multi-dimensional nature of the data set [@Kirpich2018]. 
As expected, modeling of the IP responses with canonical linear regression with Akaike information criterion-driven backward elimination (function `stepAIC()`, package MASS) [@Ripley2022] led to highly over-parameterized models as evident from huge expansion of the fit errors in cross-validation (CV) (not shown).

The Elastic Net and LASSO algorithms are implemented in R by the _glmnet_ package [@Friedman2010]. 
The optimal $\lambda$ parameters for Elastic Net and LASSO regression were found as values associated with the minimum model deviance of the optimally regularized model in 200-repeats 10-fold CV (function `cv.glmnet()`, Elastic Net: $\alpha$ = 0.5, LASSO: $\alpha$ = 1, $\lambda$ parameter 'lambda.1se' corresponding to the optimally regularized model). 
The optimal $\lambda$ values in Elastic Net were 
`r filter(eln_mod$opt_lambda, response == "ipq_total")$lambda %>% signif(2)` for the total IP score, 
`r filter(eln_mod$opt_lambda, response == "ipq_sub1")$lambda %>% signif(2)` for the emotion/concern/consequences and 
`r filter(eln_mod$opt_lambda, response == "ipq_sub2")$lambda %>% signif(2)` for the lacking control/coherence IP component scores.
The optimal $\lambda$ values in LASSO were 
`r filter(lasso_mod$opt_lambda, response == "ipq_total")$lambda %>% signif(2)` for the total IP score, 
`r filter(lasso_mod$opt_lambda, response == "ipq_sub1")$lambda %>% signif(2)` for the emotion/concern/consequences and 
`r filter(lasso_mod$opt_lambda, response == "ipq_sub2")$lambda %>% signif(2)` for the lacking control/coherence IP component score. 
Elastic Net and LASSO models with the optimal $\lambda$ values were constructed with _caret_ package (function `train(method = 'glmnet')`) [@Kuhn2008] and their performance in the training data set and 10-repeats 10-fold CV was tested.
As shown in __Supplementary Figure S\@ref(fig:fig-performance)__, for the total IP score and the emotion/concern/consequences IP component score, meaningful, reproducible and acceptably parameterized models could be built with either of the algorithms as suggested by comparison of the model performance in the training and CV data sets. 
In turn, poorly performing, estimate-only models could be obtained for the lacking control/coherence IP component. 
Estimates of non-zero coefficients were extracted from the final models with the _coef_ method (the `s` parameter set to the optimal $\lambda$ value) (__Supplementary Figures S\@ref(fig:fig-eln-est-total)__ - __S\@ref(fig:fig-blasso-est-sub1)__).

The Bayesian LASSO algorithm is implemented in R by the _monomvn_ package [@Gramacy2022]. 
The Bayesian LASSO model was constructed with the _caret_ package (function `train(method = 'blasso')`, iteration number: T = 1000). 
The optimal value of the sparsity parameter, which controls the fraction of non-zero coefficients in the final model based on the posterior distributions, was found by 10-repeats 10-fold CV with the RMSE (root mean squared error) statistic as a selection criterion. 
The optimal sparsity values selected from the 0.1 - 0.7 range were `r blass_mod$models$ipq_total$bestTune[1, "sparsity"] %>% signif(2)`, `r blass_mod$models$ipq_sub1$bestTune[1, "sparsity"] %>% signif(2)` and `r blass_mod$models$ipq_sub2$bestTune[1, "sparsity"] %>% signif(2)` for the total IP score, the emotion/concern/consequences and the lacking control/coherence IP component scores, respectively. 
Similar to the Elastic Net and LASSO algorithms, meaningful models could be obtained for the total IP score and the emotion/concern/consequences IP component score but not for the lacking control/coherence IP component score (__Supplementary Figure S\@ref(fig:fig-performance)__). 
The model coefficient matrix X was extracted from the 'finalModel' slot of the caret model. 
The coefficient values (__Supplementary Figures S\@ref(fig:fig-blasso-est-total)__ and __S\@ref(fig:fig-blasso-est-sub1)__) were calculated as matrix column medians over all algorithm iterations [@Park2012].

>>>>

set_flextable_defaults(font.family = 'Cambria', font.size = 9)

@@

The CovILD longitudinal observation cohort (ClinicalTrials.gov: NCT04416100) includes adult ($\geq$ 18 year) patients with a symptomatic, PCR-confirmed SARS-CoV-2 infection. 
The follow-up visits were scheduled at two, three, six months and one year after COVID-19 diagnosis. 
In the current report, a subset of the original CovILD cohort was investigated, which displayed (1) persistent somatic symptoms (PSS) or (2) any abnormality in chest computed tomography or (4) any deficits in lung function testing or (5) any cardiological abnormality in trans-thoracic echocardiography at the one-year follow-up. 
The final analysis inclusion criterion was availability of complete Brief Illness Perception Questionnaire (BIPQ) [@Broadbent2006] and the complete set of variables used in explorative data analysis, modeling and clustering (__Supplementary Table S\@ref(tab:tab-var-dict)__ and __S\@ref(tab:tab-modeling-vars)__).

The decision to analyze the study participant subset with COVID-19 sequlae was motivated by the design of the illness perception-measuring tool used in our report, the BIPQ, which was initially developed for and validated in somatic conditions [@Broadbent2006].

@@

The protocol of each follow-up visit included a survey of PSS, blood laboratory parameter assessment, lung function testing, trans-thoracic echocardiography and chest computed tomography as described in more detail before [@Sonnweber2020; @Sonnweber2022; @Luger2022]. 
Comorbidities and data on acute COVID-19 course (symptoms, laboratory parameters, functional tests and imaging) were recorded retrospectively at the two-month follow-up visit based on information provided by the participant and electronic patient records.

@@

The baseline demographic variables included sex (male/female), age at COVID-19 diagnosis, smoking history, body weight and height. 
The following pre-existing comorbidities were surveyed: obesity, metabolic disease, cardiovascular disease, hypertension, type II diabetes, hypercholesterolemia, gastrointestinal disease, malignancy, chronic kidney disease, immune deficiency, pulmonary and respiratory conditions. 
Among the surveyed comorbidities, asthma, chronic obstructive pulmonary disease, interstitial lung disease and other chronic pulmonary disorders were subsumed under the 'respiratory disease' variable included in the independent modeling variable set.
Study participants were stratified according to the acute COVID-19 severity as ambulatory (outpatient, WHO ordinal scale for clinical improvement 1 - 2), moderate (hospitalized at normal infection ward, no oxygen therapy, WHO 3 - 4) and severe COVID-19 survivors (hospitalized with oxygen therapy or mechanical ventilation or intensive care, WHO 5 - 7).

@@

Lung function testing abnormality was defined as 
forced vital capacity < 80% or 
forced expiratory volume in 1 second < 80% or 
total lung capacity < 80% or 
diffusion lung capacity for carbon monoxide < 80% of the predicted reference value or 
the ratio of forced expiratory volume in 1 second to forced vital capacity < 70% of the predicted reference value [@Sonnweber2022]. 
Computed tomography images were screened for ground glass opacities, consolidations, bronchiectasis and reticulations according to the Fleischner Society glossary [@Hansell2008]. 
Chest computed tomography abnormalities were scored separately for each lobe with the computed tomography severity score [@Sonnweber2020; @Sonnweber2022; @Luger2022]: 0 - no abnormality, 1 - minimal (subtle ground glass opacities), 2 - mild (several ground glass opacities, subtle reticulation), 3 - moderate (multiple ground glass opacities, reticulation, small consolidation), 4 - severe (extensive ground glass opacities, consolidation, reticulation with distortion), 5 - massive (multiple findings, parenchymal destruction). 
The sum computed tomography severity score for all five lobes was used in the analysis. 
Any chest computed tomography abnormality is defined as computed tomography severity score $\geq$ 1. 
The most frequent cardiological abnormality in echocardiography was low grade diastolic dysfunction. 
No reduced left ventricular ejection fraction was observed in the participants included in the current analysis. 
Frequencies of cardiopulmonary findings at the one-year follow-up are presented in __Supplementary Table S\@ref(tab:tab-cohort-fup)__ and __Supplementary Figure S\@ref(fig:fig-overlap)__.

@@

- __reduced physical performance__ defined as Eastern Cooperative Oncology Group (ECOG) scale $\geq$ 1
- __dyspnea__ defined as Modified Medical British Research Council (mMRC) scale $\geq$ 1
- __cough__, self-reported, surveyed as a yes/no item
- __sleep problems__, self-reported, surveyed as a yes/no item
- __night sweating__, self-reported, surveyed as a yes/no item
- __hyposmia or anosmia__, self-reported, surveyed as a yes/no item
- __dermatological symptoms__, self-reported, surveyed as a yes/no item
- __gastrointestinal symptoms__, self-reported, surveyed as a yes/no item
- __hair loss__, self-reported, surveyed as a yes/no item
- __significant fatigue__, defined as bimodal Chalder's Fatigue Scale $\geq$ 4 [@Chalder1993; @Morriss1998]

@@

Since only few participants displayed values of Eastern Cooperative Oncology Group scale $\geq$ 2 or Modified Medical British Research Council scale $\geq$ 2, the dichotomous 'reduced performance (ECOG $\geq$ 1)' and 'dyspnea (mMRC $\geq$ 1)' independent variables were used in modeling instead of the full numeric scales.
Fatigue intensity was rated with the likert Chalder's Fatigue Scale [@Chalder1993; @Morriss1998]. 
Exertional capacity was assessed by six-minute walking test performed according to the American Thoracic Society guidelines [@Crapo2012]. 
The reference six-minute walking distance values were obtained as described by Crapo and colleagues [@Crapo2012] and differences between the actual and reference values were calculated.
Frequencies of symptoms, ratings of fatigue and exertional capacity at the one-year follow-up are shown in __Supplementary Table S\@ref(tab:tab-cohort-fup)__, __Supplementary Figures S\@ref(fig:fig-sev-symptoms)__ and  __S\@ref(fig:fig-overlap)__.

@@

Laboratory parameters encompassing hemoglobin (total and glycated hemoglobin), biomarkers of iron turnover (ferritin, transferrin saturation, hepcidin and soluble transferrin receptor), inflammation and coagulation (C-reactive protein and D-dimer) and cardiovascular pathology (N-terminal pro-brain natriuretic peptide) were determined routinely at the certified laboratory of the Central Institute for Medical and Chemical Diagnostic at the University Hospital in Innsbruck. 
Other markers of systemic inflammation, interleukin 6 and procalcitonin, were within the normal ranges in the analysis collective. 
Values of laboratory parameters at the one-year follow-up are listed in __Supplementary Table S\@ref(tab:tab-cohort-biochemistry)__.

@@

Illness perception (IP) was scored with a German translation of BIPQ [@Broadbent2006]. 
Each of the 8 BIPQ items (questions Q1 - Q8) was scored in with an 11-point likert scale, the items Q3, Q4 and Q7 were conceptualized as negative items are were inverted prior to the analysis:

@@

The total IP score was calculated as an arithmetic sum of items Q1 - Q8. 
As shown by the results of factor analysis [@Bartlett1937] of the BIPQ tool items (__Supplementary Figure S\@ref(fig:fig-factor-analysis)__), two separate factors could be identified: the emotion/concern/consequences (items Q1, Q2, Q5, Q6, Q8) and the lacking control/coherence component (items Q3, Q4, Q7). 
For this reason, two IP component scores were calculated and analyzed: the emotion/concern/consequences score defined as the Q1, Q2, Q5, Q6, Q8 item sum and the lacking control/coherence score defined as the Q3, Q4, Q7 item sum. 
Values of the total IP score, the emotion/concern/consequences and lacking control/coherence IP component scores as well as values of the BIPQ items recorded at the one-year follow-up are presented in __Table 2__.

@@

For descriptive statistic, correlation and statistical testing for differences between analysis groups the package _rstatix_ [@Kassambara2021] and the development package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) were employed.
Factor analysis was accomplished with the _stats_ package of base R and the development package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
The BIPQ score consistency was investigated with the _psych_ package [@Revelle2022].

For modeling and ANOVA, the development package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc, backward elimination and two-way ANOVA) and the packages _glmnet_ (Elastic Net and LASSO) [@Friedman2010], _monomvn_ [@Gramacy2022] (Bayesian LASSO) and _caret_ [@Kuhn2008] were employed. 
Model quality control, cross-validation and fit statistics were computed with _caret_ and the development package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra).

Clustering was accomplished with tools provided by the packages _cluster_ [@Schubert2019], _philentropy_ [@Drost2018], _factoextra_ [@Kassambara2020] and the development package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

The _ggplot2_ [@Wickham2016] package were used for data visualization. 
Bar and violin plots were generated with [_ExDA_](https://github.com/PiotrTymoszuk/ExDA). 
Quasi-proportional Venn diagrams were generated with the _nVennR_ package [@Perez-Silva2018]. 
Figures were constructed with _cowplot_ [@Wilke2019]. 
Tables were generated with _flextable_ [@Gohel2022]. 
The manuscript and supplementary material files were created with the _rmarkdown_ [@Allaire2022], _knitr_ [@Xie2022], _bookdown_ [@Xie2016] packages and [_Affiliations_](https://github.com/drdanholmes/Affiliations) Pandoc filter set. 
For management of figures, tables and R code in the RMarkdown documents, the development package [_figur_](https://github.com/PiotrTymoszuk/figur) was used.

@@

Descriptive statistic: median with interquartile ranges for numeric variables, and percentages and counts for categorical variables were calculated with the `explore()` function ([_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package). 
Statistical significance of differences in frequency of categorical variable levels was assessed by $\chi^2$ test with Cramer V effect size statistic (`compare_variables()`, the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package). 
Normality was tested with Shapiro-Wilk test and quantile-quantile plots (`explore()`, [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package).

Since multiple study variables were non-normally distributed, differences in numeric variables were assessed by non-parametric Mann-Whitney (effect size: r) or Kruskal-Wallis test (effect size: $\eta^2$), as appropriate. 
Group comparisons were done with the `compare_variables()` function (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
Correlations were analyzed with Spearman test (`correlate_variables()` package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) and visualized by fitting a second order trend (`geom_smooth()`, package _ggplot_, formula: $y \sim x + I(x^2)$). 
Statistical significance for changes in frequency of PSS in time was investigated with Cochran Q test (function `cochran_qtest()`, package _rstatix_) [@Kassambara2021].

@@

BIPQ dimensionality was assessed by maximum  likelihood factor analysis [@Bartlett1937] (`reduce_data(red_fun = 'fa')`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), a wrapper around `factanal()` from _stats_) and visual investigation of the BIPQ item loadings (method `plot(type = 'loadings')` called for the factor analysis object). 
The factor analysis revealed presence of two components of the BIPQ tool: the emotion/concern/consequences and the lacking control/coherence factor (__Figure \@ref(fig:fig-factor-analysis)__).
Spearman's pair-wise correlations between the BIPQ items were calculated (function `correlate_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) and visualized in a bubble plot. 
Due to multi-dimensionality of the BIPQ tool McDonald's $\omega$ was calculated with the function `omega()` from the package _psych_ [@Revelle2022; @McDonald1999] was used as a reliability measure for the global BIPQ and the emotion/concern/consequences and lacking control/coherence factors (__Supplementary Figure S\@ref(fig:fig-coherence)__).

@@

In multi-parameter modeling, a set of `r length(mod$variables)` independent variables was used (__Supplementary Table S\@ref(tab:tab-modeling-vars)__). 
To improve normality, ferritin and soluble transferrin receptor concentrations were transformed with the logarithm function, and transferrin saturation and hepcidin were transformed with square root. 
Strongly non-normally or bimodally distributed numeric explanatory variables: 
number of comorbidities (`r levels(cov_data$clear_data$no_comorb_class) %>% paste(collapse = ', ')`), 
number of symptoms (`r levels(cov_data$clear_data$sympt_number_class) %>% paste(collapse = ', ')`), 
computed tomography severity score ('none': 0, 'mild': 1 - 5, 'moderate': 5 - 10, 'severe': 11+) [@Luger2022], 
N-terminal pro-brain natriuretic peptide (`r levels(cov_data$clear_data$NTproBNP_class) %>% paste(collapse = ', ')`), 
D-Dimer (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`), 
C-reactive protein (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`) and 
glycated hemoglobin (`r levels(cov_data$clear_data$DDimer_class) %>% paste(collapse = ', ')`) were stratified with clinically applicable cutoffs or cutoffs determined by visual analysis of scatter plots of modeling variable and the explanatory factor with fitted LOESS trends. 
For numeric explanatory variables, both first- and second-order terms were included in the model to account for possible non-linear relationship between the independent variable and response. 
During pre-processing, numeric variables (independent and responses) were normalized with mean centering (Z-scores).

@@

Three L1 regularized regression algorithms were utilized to model the total IP score and the IP component scores at one year after COVID-19: Elastic Net [@Zou2005], LASSO (least absolute shrinkage and selection operator) [@Tibshirani1996] and Bayesian LASSO [@Park2012]. 
The choice of penalized regression over canonical backward elimination or forward selection was motivated by highly multi-dimensional nature of the data set [@Kirpich2018]. 
As expected, modeling of the IP responses with canonical linear regression with Akaike information criterion-driven backward elimination (function `stepAIC()`, package MASS) [@Ripley2022] led to highly over-parameterized models as evident from huge expansion of the fit errors in cross-validation (not shown).

The Elastic Net and LASSO algorithms are implemented in R by the _glmnet_ package [@Friedman2010]. 
The optimal $\lambda$ parameters for Elastic Net and LASSO regression were found as values associated with the minimum model deviance of the optimally regularized model in 200-repeats 10-fold CV (function `cv.glmnet()`, Elastic Net: $\alpha$ = 0.5, LASSO: $\alpha$ = 1, $\lambda$ parameter 'lambda.1se' corresponding to the optimally regularized model). 
The optimal $\lambda$ values in Elastic Net were 
`r filter(eln_mod$opt_lambda, response == "ipq_total")$lambda %>% signif(2)` for the total IP score, 
`r filter(eln_mod$opt_lambda, response == "ipq_sub1")$lambda %>% signif(2)` for the emotion/concern/consequences and 
`r filter(eln_mod$opt_lambda, response == "ipq_sub2")$lambda %>% signif(2)` for the lacking control/coherence IP component scores.
The optimal $\lambda$ values in LASSO were 
`r filter(lasso_mod$opt_lambda, response == "ipq_total")$lambda %>% signif(2)` for the total IP score, 
`r filter(lasso_mod$opt_lambda, response == "ipq_sub1")$lambda %>% signif(2)` for the emotion/concern/consequences and 
`r filter(lasso_mod$opt_lambda, response == "ipq_sub2")$lambda %>% signif(2)` for the lacking control/coherence IP component score. 
Elastic Net and LASSO models with the optimal $\lambda$ values were constructed with _caret_ package (function `train(method = 'glmnet')`) [@Kuhn2008] and their performance in the training data set and 10-repeats 10-fold cross-validation was tested.
As shown in __Supplementary Figure S\@ref(fig:fig-performance)__, for the total IP score and the emotion/concern/consequences IP component score, meaningful, reproducible and acceptably parameterized models could be built with either of the algorithms as suggested by comparison of the model performance in the training and the cross-validation data sets. 
In turn, poorly performing, estimate-only models could be obtained for the lacking control/coherence IP component. 
Estimates of non-zero coefficients were extracted from the final models with the `coef()` method (the `s` parameter set to the optimal $\lambda$ value) (__Supplementary Figures S\@ref(fig:fig-eln-est-total)__ - __S\@ref(fig:fig-blasso-est-sub1)__).

The Bayesian LASSO algorithm is implemented in R by the _monomvn_ package [@Gramacy2022]. 
The Bayesian LASSO model was constructed with the _caret_ package (function `train(method = 'blasso')`, iteration number: T = 1000). 
The optimal value of the sparsity parameter, which controls the fraction of non-zero coefficients in the final model based on the posterior distributions, was found by 10-repeats 10-fold cross-validation with the root mean squared error as a selection criterion. 
The optimal sparsity values selected from the 0.1 - 0.7 range were `r blass_mod$models$ipq_total$bestTune[1, "sparsity"] %>% signif(2)`, `r blass_mod$models$ipq_sub1$bestTune[1, "sparsity"] %>% signif(2)` and `r blass_mod$models$ipq_sub2$bestTune[1, "sparsity"] %>% signif(2)` for the total IP score, the emotion/concern/consequences and the lacking control/coherence IP component scores, respectively. 
Similar to the Elastic Net and LASSO algorithms, meaningful models could be obtained for the total IP score and the emotion/concern/consequences IP component score but not for the lacking control/coherence IP component score (__Supplementary Figure S\@ref(fig:fig-performance)__). 
The model coefficient matrix X was extracted from the `finalModel` slot of the caret model. 
The coefficient values (__Supplementary Figures S\@ref(fig:fig-blasso-est-total)__ and __S\@ref(fig:fig-blasso-est-sub1)__) were calculated as matrix column medians over all algorithm iterations [@Park2012].

Extraction of caret model performance measures: root mean squared error and $R^2$ defined as $1 - mean\;squared\;error/variance(y)$ (__Supplementary Figure S\@ref(fig:fig-performance)__) was accomplished by the in-house-developed package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) (modeling objects created with `as_caretex()`). 
Model assumption testing (normality of residuals, Shapiro-Wilk test, method `residuals()`) and visual quality control based on plots of residuals versus fitted and residuals quantile-quantile plots was done with the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package (method `plot()` called for the `caretex` object).
